The abundance of Bullet-Groups in $\Lambda$CDM

### 2014-03-20 paper/ms.tex
geometry and measurement

We use the Bolshoi Run, a cosmological DM only simulation over a cubic
volume of 250\hMpc comoving on a side. The simulation uses the ART
code \citep{Kravtsov1997} to follow the evolution of a dark matter
density field sampled with $2048^3$ from $z=80$ to $z=0$. The
cosmology used  corresponds to  the spatially flat concordance model
with the following parameters:  the density parameter for matter (dark
matter and baryons) $\Omega_m=0.27$, the density parameter for baryonic matter
$\Omega_b=0.0469$, the density parameter for dark energy
$\Omega_{\Lambda}=0.73$, the Hubble parameter $h=0.7$, the
normalization of the Power spectrum $n=0.95$ and the amplitude of mass
density fluctuation (at redshift z$=$0) $\sigma_8=0.82$.  These
cosmological parameters are consistent with the nine-year Wilikinson
Microwave Anisotropy Probe (WMAP) results \citep{hinshaw_etal13}. A
detailed presentation of the simulation can be found in
\citet{2011ApJ...740..102K}.

The number of particles used for each of the DM component was
$2048^3$, resulting in a mass resolution of $1.35 \times 10^8$
M$_{\odot}$h$^{-1}$. The completeness limit in this simulation is set
for halos with $100$ particles corresponding to a mass of
$1.35\times10^{10}$\hMsun\ or a maximum circular velocity $V_{c}$ of
$50$\kms.

We use halo catalogs constructed using the Bound Density Maxima (BDM)
algorithm \citep{BDM,BDMb}. To define the radius of a halo we use a
density threshold of 360 times the mean density of the Universe (or
200 times the critical density), for different redshifts we use the
overdensity criterion by \cite{Bryan1998}. An important feature of BDM
is that it allows us to detect sub-halos inside larger virialized
structures.

@@

In this Letter we make a study at four different redshifts $z=0.0,0.25,
0.5$ and $1.0$. First, we select all the host halos (i.e. halos that
are not inside a larger halo) with circular velocities $V_{\rm c}\geq
300$\kms ($\geq 1\times 10^{13} \hMsun$). Then, we select the sub-halos with
circular velocities $V_{\rm c}\geq 75$\kms ($\geq 5\times
10^{10}\hMsun$). The objective is to use this sub-halos as the tracer
of the sub-dominant dark matter clump in the merging cluster, i.e. the
bullets.

These two sets (host halos and sub-halos)
constitute the basis for our analysis. We associate each host halo
to its most massive sub-halo, each one of these pairs (host \& sub-halo)
is considered as a potential Bullet system and is kept for the
analysis described in the next Section. The number of systems at each
redshift is listed on the first arrow of Table \ref{table:numbers}.

@@

The Bullet-like configurations are composed by two dark matter
structures: the host halo and the dominant sub-halo. We describe the
kinematics of this configuration by the 
position and velocity vectors of the sub-halo in a frame of reference
where the main halo is at rest; thus
$\vec{v{}}=\vec{v}_{sub}-\vec{v}_{halo}$ and
$\vec{r}=\vec{r}_{sub}-\vec{r}_{halo}$, where the subscripts $host$
and $sub$ refer to the position of the minium of potential for the
host and sub-halo in the frame of reference of the simulation,
respectively.

The angle between these two vectors can be characterized by, 
\begin{equation}
  \mu\equiv
  \cos(\theta)=\frac{\vec{v{}}\cdotp{}\vec{r}}{\left\|\vec{v}{}\right\|
    \left\|\vec{r}\right\|} .
 \end{equation} 
%
This encodes relevant information to describe the collision,
i.e. cases of $|\mu|\approx 1$ can be considered as head-on collisions
while $|\mu|\approx 0$ describes a grazing trajectory.

The geometrical configuration can be further described by the
following quantities. The circular velocity of each component, $V_{\rm
  c,host}$ for the host and $V_{\rm c, sub}$ for the sub-halo and the
size of the host halo $R_{\rm vir}$. Another useful quantity computed
in the simulation is the distance between the minium of potential for
the host halo and its center of mass, $X_{\rm
  off}=||\vec{r}_{min}-\vec{r}_{cm}||/R_{\rm vir}$ which serves as a
measurement of how much the host halo is perturbed.

As a zero-th order approximation, in this paper we work with three
observed quantities for Bullet-like systems. The projected distance
between two dominant DM clumps, the projected distance between the
DM-baryonic clumps and the ratio of the mass associated to the DM
clumps.

From the simulation point of view, the first quantity can be
translated into the 2D projected values of $||\vec{r}||$ and its value
relative to the virial radius $D_{\rm   off}= ||\vec{r}||_{\rm
  2D}/R_{\rm vir}$. The last quantity, the mass ratio, can be
approximated by ratio of the circular velocities of the two clumps $V_{\rm c,sub}/V_{\rm  c,host}$ which should be equal to the square root of the mass
ratio. The second quantity, the projected DM-baryon distance, is not
directly available from a DM-only simulation but, as we show in the
Results section, can be inferred from the available information.

The physical quantities described above can be used to describe
the three main stages in a bullet-like encounter. First, the sub-halo
crosses the virial radius of the host halo starting a head on
collision, $||\vec{r}||/R_{\rm vir}\approx 1$ and
$\mu\lessapprox<0.0$. Second, as the sub-halo crosses for the first time
the center of the host halo $||\vec{r}||/R_{\rm vir} < 1.0$ and
$\mu>0.0$. Third, as the sub-halo reaches apogee and comes back to the
center of the halo $||\vec{r}||/R_{\rm vir} < 1.0$ and $\mu<0.0$. We use
these quantities in the next section to fully characterize the different
kind of interactions observed in the Bolshoi Run,

@@

Figure \ref{fig:displacement} shows the results for two different
populations; groups with $300\kms < V_{\rm c, host}<700\kms$ and
clusters with $V_{\rm c,host}>700\kms$. Additionally, this is
presented for all redshifts $z=0.0, 0.25, 0.5$ and $1.0$.

The panel with the projected 2D physical displacements also shows a
vertical stripe with the estimated displacement for the Bullet-group reported by
\cite{Gastaldello}. In the group sample we see that a fraction of
$40\%$ to $60\%$  should present a displacement equal than the
estimate for \bullg; in the cluster sample this fraction increases to
$70\%$-$80\%$. The uncertainties in this estimates are derived from
the uncertainties in the displacement measurement for \bullg.This
fraction is naturally higher in more massive systems because they are
larger in size. Normalizing the displacements by the virial radius,
right panel Figure \ref{fig:displacement}, we see that the
distribution is the same regardless of the sample and the redshift.

>>>>

We use the Bolshoi Run, a cosmological DM only simulation over a cubic
volume of 250\hMpc comoving on a side. The simulation uses the ART
code \citep{Kravtsov1997} to follow the evolution of a dark matter
density field from $z=80$ to $z=0$ sampled with $2048^3$  particles. The
cosmology used  corresponds to  the spatially flat concordance model
with the following parameters:  the density parameter for matter (dark
matter and baryons) $\Omega_m=0.27$, the density parameter for baryonic matter
$\Omega_b=0.0469$, the density parameter for dark energy
$\Omega_{\Lambda}=0.73$, the Hubble parameter $h=0.7$, the
normalization of the Power spectrum $n=0.95$ and the amplitude of mass
density fluctuation (at redshift z$=$0) $\sigma_8=0.82$.  These
cosmological parameters are consistent with the nine-year Wilikinson
Microwave Anisotropy Probe (WMAP) results \citep{hinshaw_etal13}. A
detailed presentation of the simulation can be found in \citet{2011ApJ...740..102K}.

This results in a mass resolution of $1.35 \times 10^8$
M$_{\odot}$h$^{-1}$ for each computational particle. The completeness
limit in this simulation is set for halos with $100$ particles
corresponding to a mass of $1.35\times10^{10}$\hMsun\ or a maximum
circular velocity $V_{c}$ of $50$\kms.

We use DM halo catalogs constructed using the Bound Density Maxima (BDM)
algorithm \citep{BDM,BDMb}. To define the radius of a halo we use a
density threshold of 360 times the mean density of the Universe (
200 times the critical density), for different redshifts we use the
overdensity criterion by \cite{Bryan1998}. An important feature of BDM
is that it allows us to detect sub-halos inside larger virialized
structures.

@@

To construct our main halo sample we follow three
steps. Firstly, we select all the host halos (i.e. halos that are not
inside a larger halo) with circular velocities $V_{\rm c}\geq 300$\kms
($\geq 1\times 10^{13} \hMsun$). Secondly, we select the sub-halos with
circular velocities $V_{\rm c}\geq 75$\kms ($\geq 5\times
10^{10}\hMsun$). Thirdly, we associate each host halo to its most
massive sub-halo.

Each one of the pairs (host \& sub-halo) is considered as a
potential Bullet system and is kept for the analysis described in the
next Section. We use the sub-halos as the tracer of the sub-dominant
dark matter clump in the merging cluster, i.e. the bullets.

Finally, we split the sample into two  populations; groups with
$300\kms < V_{\rm c, host}<700\kms$ and clusters with $V_{\rm
  c,host}>700\kms$. These two samples are constructed at four redshifts:
$z=0.0, 0.25, 0.5$ and $1.0$.  The number of pairs at each redshift in
each population is listed on the first arrow of Table
\ref{table:numbers}.

\ref{table:numbers}.

@@

Bullet-like configurations have two dominand dark matter structures:
the host halo and the dominant sub-halo. We describe the kinematics of this configuration by the 
position and velocity vectors of the sub-halo in a frame of reference
where the main halo is at rest; thus
$\vec{v{}}=\vec{v}_{\rm sub}-\vec{v}_{\rm host}$ and
$\vec{r{}}=\vec{r}_{\rm sub}-\vec{r}_{\rm host}$, where the subscripts $host$
and $sub$ refer to the position of the minium of potential for the
host and sub-halo in the frame of reference of the simulation, respectively.

The angle between these two vectors can be quantified by, 
\begin{equation}
  \mu\equiv
  \cos(\theta)=\frac{\vec{v{}}\cdotp{}\vec{r}}{\left\|\vec{v}{}\right\|
    \left\|\vec{r}\right\|} .
 \end{equation} 
%
This encondes information about the collision, i.e. cases of
$|\mu|\approx 1$ can be considered as head-on collisions while
$|\mu|\approx 0$ describes a grazing trajectory.

The geometrical configuration can be further described by the
following quantities. The circular velocity of each component, $V_{\rm
  c,host}$ for the host and $V_{\rm c, sub}$ for the sub-halo and the
size of the host halo $R_{\rm vir}$. Another useful quantity computed
in the simulation is the distance between the minimum of potential for
the host halo and its center of mass (computed from all the particles
inside the $R_{\rm vir}$), $X_{\rm off}=||\vec{r}_{min}-\vec{r}_{cm}||/R_{\rm vir}$, which serves as a
measurement of how much the host halo is perturbed.

In this letter we work with three quantities that could be inferred
from observations of Bullet-like systems. The projected distance
between two dominant DM clumps, $d_{\rm 2D}$; the projected distance
between the DM-baryonic clumps, $d_{\rm 2D}^{\rm bar}$; and the mass
ration associated to the two dominant DM clumps.

From the simulation point of view, the first quantity can be
translated into the 2D projected values of $||\vec{r}||$ and its value
relative to the virial radius $D_{\rm   off}= ||\vec{r}||_{\rm
  2D}/R_{\rm vir}$. The second quantity, the projected DM-baryon
distance, is not directly available from a DM-only simulation but can
be estimated from the data.  The last quantity, the mass ratio, can be
approximated by ratio of the circular velocities of the two clumps
$V_{\rm c,sub}/V_{\rm  c,host}$ which should be equal to the square
root of the mass ratio.

We also use the physical quantities described above to discriminate
between grazing trajectories, $|\mu|<0.9$, and Bullet-like collissions
with $|\mu|\approx 1$. In the latter case we can also distinguish three
main stages in a such encounter. First, when the sub-halo crosses the
virial radius of the host halo starting a head on collision, $D_{\rm
  off} \approx 1$ and $\mu\approx -1$. Second, as the sub-halo crosses
for the first time the center of the host halo $D_{\rm off}<1.0$ and
$\mu\approx 1$. Third, as the sub-halo reaches apogee and comes back
to the center of the halo $D_{\rm off} < 1.0$ and $\mu< -1$.

@@

Figure \ref{fig:displacement} shows the results for two different
populations at all redshifts. The panel with the projected 2D physical displacements also shows a
vertical stripe with the estimated displacement for the Bullet-group reported by
\cite{Gastaldello}. In the group sample we see that a fraction of
$40\%$ to $60\%$  should present a displacement equal than the
estimate for \bullg; in the cluster sample this fraction increases to
$70\%$-$80\%$. The uncertainties in this estimates are derived from
the uncertainties in the displacement measurement for \bullg.This
fraction is naturally higher in more massive systems because they are
larger in size. Normalizing the displacements by the virial radius,
right panel Figure \ref{fig:displacement}, we see that the
distribution is the same regardless of the sample and the redshift.
